---
title: "Getting Started with ggswim"
output: rmarkdown::html_vignette
description: >
  Start here for an introduction on how to use ggswim to make swimmer plots.
vignette: >
  %\VignetteIndexEntry{Getting Started with ggswim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dev = "ragg_png",
  dpi = 400,
  fig.width = 12, fig.height = 6
)
```

The ggswim package helps you build swimmer plots in R with ggplot2-style layers. This vignette walks through the main workflow: adding patient timelines, placing events on those timelines, styling marker legends, and applying final plot polish.

ggswim offers several functions that fit into familiar ggplot2 workflows:

- `geom_swim_lane()`
- `geom_swim_marker()`
- `scale_marker_discrete()`
- `geom_swim_arrow()`
- `scale_arrow_discrete()`
- `theme_ggswim()`

`geom_swim_lane()`, an extension of `geom_segment()`, constructs the horizontal timelines, which we'll refer to as "lanes." `geom_swim_marker()` wraps `geom_text()` to place key events of interest, or "markers," on those lanes. These markers can take the form of shapes, symbols, or emojis. Arrows and themes are covered briefly here, with more examples in the focused articles.

Drawing from the well-established principles of ggplot2, ggswim lets users apply familiar layer-building techniques, including scales, labels, and themes.

### Adding a Lane Layer

Let's get started building our swimmer plot! As with the README, we will be using ggswim's internal datasets: `patient_data`, `infusion_events`, and `end_study_events`. 

Let's start with observing `patient_data`'s structure:

```{r patient_data , echo=FALSE, include=TRUE}
library(ggswim)
library(ggplot2)

patient_data
```

`patient_data` contains a long dataset where patient IDs (`pt_id`) can be repeated. These rows are differentiated by `disease_assessment` values combined with corresponding start and end times, measured in months. Together, these rows detail clinical trial timelines for each patient.

- `disease_assessment` is broken down into a few categories where:
   - CR = "Complete Response"
   - CRi = "Complete Response with Incomplete Blood Count Recovery"
   - RD = "Relapsed Disease"

Now, let's make the plot using `ggplot()` and our first layer with `geom_swim_lane()`:

```{r ggswim plot, fig.alt="Initial swimmer plot with lanes."}
library(ggplot2)

p <- ggplot() +
  geom_swim_lane(
    data = patient_data,
    mapping = aes(
      x = start_time,
      xend = end_time,
      y = pt_id,
      color = disease_assessment
    ),
    linewidth = 3
  )

p
```

Here we have patient timelines split into disease assessment intervals. `geom_swim_lane()` does the work of setting up `geom_segment()` and readying our plot layers for additional ggswim-specific features such as the "markers" mentioned earlier. It's worth noting that `geom_swim_lane()` is a thin wrapper around `geom_segment()` and supports the same functionality apart from a `yend` since swimmer plots tend to be horizontal.

## Adding a Marker Layer: Points

Now, let's add a marker layer by first inspecting the `infusion_events` and `end_study_events` datasets:

```{r infusion_events}
infusion_events
```

This dataset is much simpler, indicating the time from the initial infusion (0), and reinfusions at some point beyond 0 (if a patient had any). These are also categorized under `label`. `glyph` and `colour` will serve as helpful specifiers when we add our markers onto our swimmer plot. Initial infusions aren't a part of this dataset since the plot itself is centered around them, and having markers for all segments at month 0 would be relatively meaningless.


Next, let's look at our end of study events, i.e. events that indicate a patient has left the study for various reasons.

```{r end_study_events}
end_study_events
```

You'll notice that this dataset includes emojis under `glyph`. In addition to shapes and symbols, ggswim supports emojis when rendering swimmer plots. If emojis or custom shapes do not render as expected, check your graphics device settings; AGG devices usually work well.

While it's common to encounter these as separate datasets in the wild, it will make our lives much easier to combine `end_study_events` and `infusion_events` together since they share roughly the same data structure and the markers exist on the same timeline.

```{r combine events, message=FALSE}
all_events <- dplyr::bind_rows(
  infusion_events,
  end_study_events
)

all_events
```

Let's now call `geom_swim_marker()` and add the events onto our plot. To do this, we will use `geom_swim_marker()`'s custom `marker` `aes()` parameter:

```{r second marker plot, message=FALSE, warning=FALSE, fig.alt="Updated swimmer plot with markers."}
p <- p +
  geom_swim_marker(
    data = all_events,
    aes(
      x = time_from_initial_infusion,
      y = pt_id,
      marker = label
    ),
    size = 4
  )

p
```

We've made a swimmer plot with lanes and two kinds of events. Notice how even though both the lanes and markers use colour internally, they are separated in the legend output. Let's take it one step further and make use of the `glyph` and `colour` columns we specified.

## A Sense of Scale

Scales are a unique component of the ggplot2 framework. For more information on them and understanding how they work, you are encouraged to read the "Scales" chapter from [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/scales). In short, scales are responsible for connecting data with aesthetics and communicating those connections through elements like the plot legend.

`scale_marker_discrete()` makes it easy to control marker appearance in the plot and legend. This is helpful because plain `geom_text()` can show event labels, but it does not give ggswim's separate marker legend behavior:

```{r, fig.alt="Example display of markers without scale_marker_discrete, leading to incorrect output."}
ggplot() +
  geom_text(
    data = all_events,
    aes(x = time_from_initial_infusion, y = pt_id, label = label, colour = glyph),
    size = 4
  )
```

`geom_text()` is useful, and it is what `geom_swim_marker()` wraps, but it does not connect event labels, display glyphs, and legend entries the way swimmer plots often need.

Thanks to ggswim we can specify what belongs in the glyph versus text elements of our legend. Additionally, we are able to use the "colour" scales under the `marker` definition so ggswim can use the same colour scale under the hood, but keep the identity of "markers" and "lanes" separate in the output:

```{r, fig.alt="Updated swimmer plot with correctly used scale_marker_discrete for marker output."}
p <- p +
  scale_marker_discrete(
    name = "Study Events",
    glyphs = all_events$glyph,
    colours = all_events$colour,
    limits = all_events$label
  )

p
```


## A Full Swimmer Plot

As mentioned in the README, all of the same ggplot2 techniques you would apply to your usual plots apply here. Below, we update the lanes to match a nicer palette with a new name and add on some plot labels:

```{r mapping finalizing, message=FALSE, warning=FALSE, fig.alt="Updated swimmer plot with updated color scales."}
library(ggplot2)

p <- p +
  theme_minimal() +
  scale_color_brewer(
    name = "Disease Assessments",
    palette = "Set1"
  ) +
  labs(title = "My Swimmer Plot") +
  xlab("Time (Months)") + ylab("Patient ID")

p
```

We can also apply `theme_ggswim()` for a swimmer-plot-oriented theme:

```{r mapping with theme_ggswim, message=FALSE, warning=FALSE, fig.alt="Updated swimmer plot with theming."}
p +
  theme_ggswim()
```

### Additional notes

Some additional considerations to keep in mind when working with ggswim:

- **Rendering Emojis and Custom Shapes:** To ensure emojis and other custom shapes display correctly, users may need to [switch their graphics rendering device](https://support.posit.co/hc/en-us/articles/360017886674-Troubleshooting-RStudio-Rendering-Errors) to AGG.
- ggswim supports use of FontAwesome and Bootstrap icons for glyph definition in addition to shapes and emojis. Check out the Gallery for examples on how to add them to your plots!
- See the Adding arrows article for a focused guide to continuation indicators with `geom_swim_arrow()` and `scale_arrow_discrete()`.
