---
title: "Getting Started with ggswim"
output: rmarkdown::html_vignette
description: >
  Start here for an introduction on how to use ggswim to make swimmer plots.
vignette: >
  %\VignetteIndexEntry{Getting Started with ggswim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dev = "ragg_png",
  dpi = 400,
  fig.width = 12, fig.height = 6
)
```

The ggswim package eases the development of swimmer plots in R through extension of ggplot2. In this vignette, we'll walk through how users can create visually striking swimmer plots.

ggswim offers several layering functions that mimic other "geom" functions from ggplot2: 

- `geom_swim_lane()`
- `geom_swim_marker()`
- `scale_marker_discrete()`

The former, an extension of `geom_segment()`, allows users to construct the horizontal bars, what we'll sometimes refer to as "lanes." Meanwhile, `geom_swim_marker()` wraps `geom_text()` to effortlessly embed key events of interest, or "markers," onto the lanes. These can take the form of shapes, symbols, or even emojis.

Drawing from the well-established principles of ggplot2, ggswim allows users to apply familiar layer-building techniques, including the application of styles and themes. You'll also see that common behaviors are still accessible with ggswim such as auto-completing `aes()` parameters with piping.

### Adding a Lane Layer

Let's get started building our swimmer plot! As with the README, we will be using ggswim's internal datasets: `patient_data`, `infusion_events`, and `end_study_events`. 

Let's start with observing `patient_data`'s structure:

```{r patient_data , echo=FALSE, include=TRUE}
library(ggswim)
library(ggplot2)

patient_data
```

`patient_data` contains a long dataset where patient ID's (`pt_id`) can be repeated. These rows are differentiated by `disease_assessments`s combined with corresponding start and end times, representing months. Together, these rows detail clinical trial timelines for a given patient.

- `disease_assessment` is broken down into a few categories where:
   - CR = "Complete Response"
   - CRi = "Complete Response with Incomplete Blood Count Recovery"
   - RD = "Relapsed Disease"

Now, let's make the plot using `ggplot()` and our first layer with `geom_swim_lane()`:

```{r ggswim plot}
library(ggplot2)

p <- ggplot() +
  geom_swim_lane(
    data = patient_data,
    mapping = aes(
      x = start_time,
      xend = end_time,
      y = pt_id,
      color = disease_assessment
    ),
    linewidth = 3
  )

p
```

Here we have a simple bar graph showing infusions grouped by patients with a given disease assessment status. `geom_swim_lane()` does the work of setting up `geom_segment()` and readying our plot layers for additional ggswim-specific features such as the "markers" mentioned earlier. It's worth noting that `geom_swim_lane()` is a very thin wrapper around `geom_segment()` and supports the same functionality apart from a `yend` since swimmer plots tend to be horizontal.

## Adding a Marker Layer: Points

Now, let's add a marker layer by first inspecting the `infusion_events` and `end_study_events` datasets:

```{r infusion_events}
infusion_events
```

This dataset is much simpler, indicating the time from the initial infusion (0), and reinfusions at some point beyond 0 (if a patient had any). These are also categorized under `label`. `glyph` and `colour` will serve as helpful specifiers when we add our markers onto our swimmer plot. Initial infusions aren't a part of this dataset since the plot itself is centered around them, and having markers for all segments at month 0 would be relatively meaningless.


Next, let's look at our end of study events, i.e. events that indicate a patient has left the study for various reasons.

```{r end_study_events}
end_study_events
```

You'll notice that this dataset includes use of emojis under `glyph`. In addition to shapes and symbols, ggswim supports the use of emojis when rendering swimmer plots. <span style="color:red">If issues arise in rendering, you may need to check your options settings and ensure a proper graphics device like "AGG".</span>

While it's common to encounter these as separate datasets in the wild, it will make our lives much easier to combine `end_study_events` and `infusion_events` together since they share roughly the same data structure and the markers exist on the same timeline.

```{r combine events, message=FALSE}
all_events <- dplyr::bind_rows(
  infusion_events,
  end_study_events
)

all_events
```

Let's now call `geom_swim_marker()` and add the events onto our plot. To do this, we will use `geom_swim_marker()`'s custom `marker` `aes()` parameter:

```{r second marker plot, message=FALSE, warning=FALSE}
p <- p +
  geom_swim_marker(
    data = all_events,
    aes(
      x = time_from_initial_infusion,
      y = pt_id,
      marker = label
    ),
    size = 4
  )

p
```

We've successfully made a swimmer plot with lanes and two different kinds of marker layers! Notice how even though both the lanes and markers use the color scale, they are separated in the legend output. Let's take it one step further and make use of the `glyph` and `colour` columns we specified.

## A Sense of Scale

Scales are a unique component of the ggplot2 framework. For more information on them and understanding how they work, you are encouraged to read the "Scales" chapter from [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/scales). In short, scales are responsible for connecting data with aesthetics and communicating those connections through elements like the plot legend.

`scale_marker_discrete()` makes it easy to specify and take finer control over your markers and their appearance in the legend. Typically, it can be difficult to have emojis and labels appear in the place of legend glyphs. Let's take a look at a typical output using base ggplot2 with `geom_text()`:

```{r}
ggplot() +
  geom_text(
    data = all_events,
    aes(x = time_from_initial_infusion, y = pt_id, label = label, colour = glyph),
    size = 4
  )
```

`geom_text()` does a decent job, and is actually what `geom_swim_marker()` wraps. However, at it's base level it doesn't quite set up the use of glyphs the way we would like to see them in the legend display. Additionally, those glyphs don't appear in the plot. You could try to rework this with `geom_point()` or `geom_label()`, but would still run into issues with scale assignment. This pain point is one of the foundational reasons we made ggswim in the first place!

Thanks to ggswim we can specify what belongs in the glyph versus text elements of our legend. Additionally, we are able to use the "colour" scales under the `marker` definition so ggswim can use the same colour scale under the hood, but keep the identity of "markers" and "lanes" separate in the output:

```{r}
p <- p +
  scale_marker_discrete(
    name = "Study Events",
    glyphs = all_events$glyph,
    colours = all_events$colour,
    limits = all_events$label
  )

p
```


## A Full Swimmer Plot

As mentioned in the README, all of the same ggplot2 techniques you would apply to your usual plots apply here. Below, we update the lanes to match a nicer palette with a new name and add on some plot labels:

```{r mapping finalizing, message=FALSE, warning=FALSE}
library(ggplot2)

p <- p +
  theme_minimal() +
  scale_color_brewer(
    name = "Disease Assessments",
    palette = "Set1"
  ) +
  labs(title = "My Swimmer Plot") +
  xlab("Time (Months)") + ylab("Patient ID")

p
```

We can also apply the `theme_ggswim()` function to give it some additional beautification:

```{r mapping with theme_ggswim, message=FALSE, warning=FALSE}
p +
  theme_ggswim()
```

### Additional notes

Some additional considerations to keep in mind when working with ggswim:

- **Rendering Emojis and Custom Shapes:** To ensure emojis and other custom shapes display correctly, users may need to [switch their graphics rendering device](https://support.posit.co/hc/en-us/articles/360017886674-Troubleshooting-RStudio-Rendering-Errors) to AGG.
