---
title: "Getting Started with ggswim"
output: rmarkdown::html_vignette
description: >
  Start here for an introduction on how to use ggswim to make swimmer plots.
vignette: >
  %\VignetteIndexEntry{Getting Started with ggswim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dev = "ragg_png",
  dpi = 400,
  fig.width = 12, fig.height = 6
)
```

The ggswim package eases the development of swimmer plots in R through extension of ggplot2 in combination with ggnewscale. In this vignette, we'll walk through how users can create visually striking swimmer plots.

ggswim offers several layering functions that mimic other "geom" functions from ggplot2: 

- `geom_swim_lane()`
- `geom_swim_marker()`
- `scale_marker_discrete()`

The former, an extension of `geom_segment()`, allows users to construct the horizontal bars, what we'll sometimes refer to as "lanes." Meanwhile, `geom_swim_marker()` wraps `geom_text()` to effortlessly embed key events of interest, or "markers," onto the lanes. These can take the form of shapes, symbols, or even emojis.

Drawing from the well-established principles of ggplot2, ggswim allows users to apply familiar layer-building techniques, including the application of styles and themes.

### Adding a Lane Layer

Let's get started building our swimmer plot! As with the README, we will be using ggswim's internal datasets: `patient_data`, `infusion_events`, and `end_study_events`. 

Let's start with observing `patient_data`'s structure:

```{r patient_data , echo=FALSE, include=TRUE}
library(ggswim)
library(ggplot2)

patient_data |>
  rmarkdown::paged_table()
```

`patient_data` contains a long pivoted dataset where patient ID's (`pt_id`) can be repeated. These rows are differentiated by `disease_assessments`s combined with corresponding start and end times, representing months. Together, these rows signify a patient survival timeline.

- `disease_assessment` is broken down by the following nomenclature along with some indicator of B-cell status (if applicable):
   - CR = "Complete Response"
   - CRi = "Complete Response with Incomplete Blood Count Recovery"
   - RD = "Relapsed Disease"

Now, let's make the plot using `ggplot()` and `geom_swim_lane()`:

```{r ggswim plot}
library(ggplot2)

p <- ggplot() +
  geom_swim_lane(
    data = patient_data,
    mapping = aes(
      x = start_time,
      xend = end_time,
      y = pt_id,
      color = disease_assessment
    ),
    linewidth = 3
  )

p
```

Here we have a simple line graph, showing infusions grouped by patients with a given disease assessment status. `geom_swim_lane()` does the work of setting up a `geom_segment()` and readying our plot layers for additional ggswim-specific features such as the "markers" mentioned earlier. 

## Adding a Marker Layer: Points

Let's add a marker layer next from the `infusion_events` dataset:

```{r infusion_events}
infusion_events |>
  rmarkdown::paged_table()
```

This dataset is much simpler, indicating the time from the initial infusion, where 0 is an initial infusion and reinfusions at some point beyond 0 (if a patient had any). These are also categorized under `label`. `glyph` and `colour` will serve as helpful specifiers when we add our markers onto our swimmer plot.


Next, let's look at our end of study events, i.e. events that indicate a patient has left the study for various reasons.

```{r end_study_events}
end_study_events |>
  rmarkdown::paged_table()
```

You'll notice that this dataset includes use of emojis under `glyph`. In addition to shapes and symbols, ggswim supports the use of emojis when rendering swimmer plots. If issues arise in rendering, you may need to check your options settings and ensure a proper graphics device like "AGG". 

While it's common to encounter these as separate datasets in the wild, it will make our lives much easier to combine `end_study_events` and `infusion_events` together since it's relatively straightforward to do and the markers exist on the same timeline.

```{r combine events, message=FALSE}
all_events <- dplyr::bind_rows(
  infusion_events,
  end_study_events
)

all_events
```

Let's now call `geom_swim_marker()` and add the events onto our plot:

```{r second marker plot, message=FALSE, warning=FALSE}
p <- p +
  geom_swim_marker(
    data = all_events,
    aes(
      x = time_from_initial_infusion,
      y = pt_id,
      marker = label
    ),
    size = 4
  )

p
```

We've successfully made a swimmer plot with lanes and two different kinds of marker layers. However, let's take it one step further and make use of the `glyph` and `colour` columns we specified. After all, who doesn't love some emojis?

## A Sense of Scale

`scale_marker_discrete()` makes it easy to specify and take finer control over your markers and their appearance in the legend. Typically, it can be difficult to have emojis or labels appear in the place of legend glyphs, but thanks to ggswim we can specify what belongs in the glyph versus text elements of our legend. Additionally, we are able to use the "colour" scales under the `marker` definition so ggswim can use the same colour scale under the hood, but keep the identity of "markers" and "lanes" separate.

```{r}
p <- p +
  scale_marker_discrete(glyphs = all_events$glyph, colors = all_events$colour, limits = all_events$label)

p
```


## A Full Swimmer Plot

Let's fix up this plot so it looks a bit prettier:

```{r mapping finalizing, message=FALSE, warning=FALSE}
library(ggplot2)

p <- p +
  theme_minimal() +
  scale_color_brewer(
    name = "Study Events",
    palette = "Set1"
  ) +
  labs(title = "My Swimmer Plot") +
  xlab("Time (Months)") + ylab("Patient ID")

p
```

We can also apply the `theme_ggswim()` function:

```{r mapping with theme_ggswim, message=FALSE, warning=FALSE}
p +
  theme_ggswim()
```

### Additional notes

Some additional considerations to keep in mind when working with ggswim:

- **Rendering Emojis and Custom Shapes:** To ensure emojis and other custom shapes display correctly, users may need to [switch their graphics rendering device](https://support.posit.co/hc/en-us/articles/360017886674-Troubleshooting-RStudio-Rendering-Errors) to AGG.
