---
title: "Adding arrows to ggswim"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 6
)
```

It can be useful to visually indicate the continuation of specific records in a swimmer plot. Optionally adding arrows to the tail ends of swimmer plot lanes in ggswim can help communicate subject survival status, such as whether or not a subject is still on a given study.

To facilitate the addition of arrows, we provide the `add_arrows()` function as a way to tack on these continuation indicators to your swimmer plot lanes. Behind the scenes, the inclusion of arrows is facilitated by a call to `ggplot2::geom_segment()` and creates a new ggplot2 layer with a `ggswim_arrows` flag. `add_arrows()` sets the segment length to 0, so that assigned arrows are always placed on the right side of indicated subjects. `add_arrows()` gives users have full control over arrow neck and head length, along with options for color, fill, and type (refer to `?add_arrows` for detailed information).

## Adding arrows using `add_arrows()`

To demonstrate how you might add arrows onto the `patient_data` dataset, let's take a subset of `patient_data` that would help us make use of `add_arrows()`:

```{r, include=TRUE}
library(ggswim)
library(ggplot2)

arrow_data <- patient_data |>
  dplyr::left_join(
    end_study_events |>
      dplyr::select(pt_id, end_study_name),
    by = "pt_id"
  ) |>
  dplyr::select(pt_id, end_time, end_study_name) |>
  dplyr::filter(.by = pt_id, end_time == max(end_time)) |>
  unique()

arrow_data |>
  rmarkdown::paged_table()
```

This should look familiar, a pared-down dataset with a `status` column that we'll use to flag which lanes of our swimmer plot should have arrows. Now let's use `add_arrows()` in combination with `ggswim()` to make a swimmer plot:

```{r}
patient_data |>
  ggplot() +
  geom_swim_lane(
    mapping = aes(
      x = start_time, xend = end_time, y = pt_id,
      color = disease_assessment
    ),
    linewidth = 5
  ) +
  add_arrows(
    data = arrow_data,
    mapping = aes(xend = end_time, y = pt_id),
    arrow_neck_length = 5,
    arrow_colour = "slateblue",
    arrow_fill = "cyan"
  )
```
