---
title: "Adding arrows to ggswim"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 6
)
```

It can be useful to visually indicate the continuation of specific records in a swimmer plot. Optionally adding arrows to the tail ends of swimmer plot lanes in ggswim can help communicate subject survival status, such as whether or not a subject is still on a given study.

To facilitate the addition of arrows, we provide the geom_swim_arrow() function as a way to tack on these continuation indicators to your swimmer plot lanes. Behind the scenes, the inclusion of arrows is facilitated by a call to `ggplot2::geom_segment()`, setting zero-length segments so that assigned arrows are always placed on the right side of indicated lanes. `geom_swim_arrow()` gives users control over arrow neck and head length, along with options for color, fill, and type (refer to `?geom_swim_arrow` for detailed information).

Arrow appearance can be specified in two ways:

- directly via parameters such as `arrow_colour`, `arrow_fill`, and `arrow_type`
- by mapping the arrow aesthetic and supplying styles through `scale_arrow_discrete()`

`geom_swim_arrow()` does not replace the ability to separately call arrows using the arrow parameter in `geom_swim_lane()`. As mentioned, `geom_swim_lane()` comes with nearly all of the same capabilities as `geom_segment()`, but it may be more challenging to apply arrows as expected depending on how your data is structured.

## Adding arrows using `geom_swim_arrow()`

To demonstrate how you might add arrows onto the `patient_data` dataset, let's take a subset of `patient_data` that would help us make use of `geom_swim_arrow()`:

```{r, include=TRUE}
library(ggswim)
library(ggplot2)

arrow_data <- patient_data |>
  dplyr::left_join(
    end_study_events |>
      dplyr::select(pt_id, label),
    by = "pt_id"
  ) |>
  dplyr::select(pt_id, end_time, label) |>
  dplyr::filter(.by = pt_id, end_time == max(end_time)) |>
  dplyr::filter(is.na(label)) |>
  unique()

arrow_data
```

This should look familiar as a pared-down subset of `end_study_events` without an indicated `label`. Since filled out label statuses from `end_study_events` dataset mean a subject went off study, arrows are only applicable for subjects with no end study status. Now let's use `geom_swim_arrow()` in combination with `geom_swim_lane()` to make a swimmer plot:

```{r, fig.alt="Swimmer plot display with arrow specification."}
patient_data |>
  ggplot() +
  geom_swim_lane(
    mapping = aes(
      x = start_time, xend = end_time, y = pt_id,
      color = disease_assessment
    ),
    linewidth = 5
  ) +
  geom_swim_arrow(
    data = arrow_data,
    mapping = aes(xend = end_time, y = pt_id),
    arrow_neck_length = 5,
    arrow_head_length = grid::unit(0.15, "inches"),
    arrow_colour = "firebrick",
    arrow_fill = "gold"
  )
```

Here we've correctly assigned arrows to only the subset of patients that have not met an end of study event. Note that `geom_swim_arrow()` gives plenty of control to the arrow head and neck color, shape, and even length.

`geom_swim_arrow()` also allows for aesthetically mapped x "start" and `xend` values. If you would prefer to have dynamically mapped arrow neck lengths indicative of your data, simply supply the `x` and `xend` values in `aes()` for `geom_swim_arrow()` to use.

## Using `scale_arrow_discrete()` for arrow styling

In addition to specifying arrow appearance directly through parameters, arrows can also be styled using the arrow aesthetic together with `scale_arrow_discrete()`. This approach allows arrow styles to be controlled through the ggplot2 scaling system, making it easier to add arrows as their own legend component or to manage multiple arrow styles.

To demonstrate this approach, we can map a value to the arrow aesthetic and then define its appearance using `scale_arrow_discrete()`:

```{r, fig.alt="Swimmer plot display with arrow specification using discrete scale."}
patient_data |>
  ggplot() +
  geom_swim_lane(
    mapping = aes(
      x = start_time, xend = end_time, y = pt_id,
      color = disease_assessment
    ),
    linewidth = 5
  ) +
  geom_swim_arrow(
    data = arrow_data,
    mapping = aes(
      xend = end_time,
      y = pt_id,
      arrow = "Continuation"
    ),
    arrow_neck_length = 5,
    arrow_head_length = grid::unit(0.15, "inches")
  ) +
  scale_arrow_discrete(
    limits = "Continuation",
    colours = "firebrick",
    fills = "gold",
    types = "closed",
    name = NULL
  )
```
  
Here, the "Continuation" value mapped to the arrow aesthetic is used by `scale_arrow_discrete()` to determine the arrow’s visual properties. Because this mapping is treated like any other discrete scale in ggplot2, it can also be displayed as a separate legend entry if desired.

Using `scale_arrow_discrete()` is particularly helpful when:

- arrows should appear as a dedicated legend component
- multiple arrow styles are required
- arrow styling should be managed consistently across plots

For simple cases, specifying arrow parameters directly in `geom_swim_arrow()` may still be sufficient. However, using `scale_arrow_discrete()` provides a more flexible and scalable approach when working with more complex swimmer plots.
