---
title: "Adding arrows to ggswim"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 6
)
```

It can be useful to visually indicate the continuation of specific records in a swimmer plot. Optionally adding arrows to the tail ends of swimmer plot lanes in ggswim can help communicate subject survival status, such as whether or not a subject is still on a given study.

To facilitate the addition of arrows, we provide the `geom_swim_arrow()` function as a way to tack on these continuation indicators to your swimmer plot lanes. Behind the scenes, the inclusion of arrows is facilitated by a call to `ggplot2::geom_segment()`, setting 0 length segments so that assigned arrows are always placed on the right side of indicated lanes. `geom_swim_arrow()` gives users full control over arrow neck and head length, along with options for color, fill, and type (refer to `?geom_swim_arrow` for detailed information).

`geom_swim_arrow()` does not replace the ability to separately call arrows using the `arrow` parameter in `geom_swim_lane()`. As mentioned, `geom_swim_lane()` comes with nearly all of the same capabilities as `geom_segment()`, but it may be more challenging to apply arrows as expected depending on how your data is structured.

## Adding arrows using `geom_swim_arrow()`

To demonstrate how you might add arrows onto the `patient_data` dataset, let's take a subset of `patient_data` that would help us make use of `geom_swim_arrow()`:

```{r, include=TRUE}
library(ggswim)
library(ggplot2)

arrow_data <- patient_data |>
  dplyr::left_join(
    end_study_events |>
      dplyr::select(pt_id, label),
    by = "pt_id"
  ) |>
  dplyr::select(pt_id, end_time, label) |>
  dplyr::filter(.by = pt_id, end_time == max(end_time)) |>
  dplyr::filter(is.na(label)) |>
  unique()

arrow_data
```

This should look familiar as a pared-down subset of `end_study_events` without an indicated `label`. Since filled out label statuses from `end_study_events` dataset mean a subject went off study, arrows are only applicable for subjects with no end study status. Now let's use `geom_swim_arrow()` in combination with `geom_swim_lane()` to make a swimmer plot:

```{r, fig.alt="Swimmer plot display with arrow specification."}
patient_data |>
  ggplot() +
  geom_swim_lane(
    mapping = aes(
      x = start_time, xend = end_time, y = pt_id,
      color = disease_assessment
    ),
    linewidth = 5
  ) +
  geom_swim_arrow(
    data = arrow_data,
    mapping = aes(xend = end_time, y = pt_id),
    arrow_neck_length = 5,
    arrow_colour = "firebrick",
    arrow_fill = "gold"
  )
```

Here we've correctly assigned arrows to only the subset of patients that have not met an end of study event. Note that `geom_swim_arrow()` gives plenty of control to the arrow head and neck color, shape, and even length.
