---
title: "Adding arrows to ggswim"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 6
)
```

It can be useful to visually indicate the continuation of specific records in a swimmer plot. Adding arrows to the tail ends of swimmer plot lanes in ggswim can help communicate subject survival status, such as whether or not a subject is still on a given study. Use this article when the README example is not enough and you need more control over arrow placement, styling, or legends.

To facilitate the addition of arrows, we provide `geom_swim_arrow()` as a way to attach continuation indicators to swimmer plot lanes. Behind the scenes, arrows are drawn with `ggplot2::geom_segment()`. When only `xend` is mapped, `geom_swim_arrow()` treats `xend` as the lane endpoint and extends a fixed-length arrow neck to the right. When both `x` and `xend` are mapped, the arrow extent is data-driven. `geom_swim_arrow()` gives users control over arrow neck and head length, along with options for colour/color, fill, and type; refer to `?geom_swim_arrow` for detailed information.

Arrow appearance can be specified in two ways:

- directly via parameters such as `arrow_colour`, `arrow_fill`, and `arrow_type`
- by mapping the arrow aesthetic and supplying styles through `scale_arrow_discrete()`

`geom_swim_arrow()` does not replace the ability to use the `arrow` parameter in `geom_swim_lane()`. As mentioned, `geom_swim_lane()` comes with nearly all of the same capabilities as `geom_segment()`, but it may be more challenging to apply arrows as expected depending on how your data is structured.

## Adding arrows using `geom_swim_arrow()`

To demonstrate how you might add arrows onto the `patient_data` dataset, let's take a subset of `patient_data` that would help us make use of `geom_swim_arrow()`:

```{r, include=TRUE}
library(ggswim)
library(ggplot2)

arrow_data <- patient_data |>
  dplyr::left_join(
    end_study_events |>
      dplyr::select(pt_id, label),
    by = "pt_id"
  ) |>
  dplyr::select(pt_id, end_time, label) |>
  dplyr::filter(.by = pt_id, end_time == max(end_time)) |>
  dplyr::filter(is.na(label)) |>
  unique()

arrow_data
```

This dataset is derived from `patient_data` joined to end-study labels. Since populated labels in `end_study_events` mean a subject went off study, arrows are only applicable for subjects with no end-study status. Now let's use `geom_swim_arrow()` in combination with `geom_swim_lane()` to make a swimmer plot:

```{r, fig.alt="Swimmer plot display with arrow specification."}
patient_data |>
  ggplot() +
  geom_swim_lane(
    mapping = aes(
      x = start_time, xend = end_time, y = pt_id,
      color = disease_assessment
    ),
    linewidth = 5
  ) +
  geom_swim_arrow(
    data = arrow_data,
    mapping = aes(xend = end_time, y = pt_id),
    arrow_neck_length = 5,
    arrow_head_length = grid::unit(0.15, "inches"),
    arrow_colour = "firebrick",
    arrow_fill = "gold"
  )
```

Here we've assigned arrows to only the subset of patients that have not met an end-of-study event. Note that `geom_swim_arrow()` gives control over the arrow head, neck, colour/color, shape, and length.

`geom_swim_arrow()` also allows aesthetically mapped `x` start and `xend` values. If you would prefer arrow neck lengths to come from your data, supply both `x` and `xend` in `aes()`.

## Using `scale_arrow_discrete()` for arrow styling

In addition to specifying arrow appearance directly through parameters, arrows can also be styled using the arrow aesthetic together with `scale_arrow_discrete()`. This approach allows arrow styles to be controlled through the ggplot2 scaling system, making it easier to add arrows as their own legend component or to manage multiple arrow styles.

To demonstrate this approach, we can map a value to the arrow aesthetic and then define its appearance using `scale_arrow_discrete()`:

```{r, fig.alt="Swimmer plot display with arrow specification using discrete scale."}
patient_data |>
  ggplot() +
  geom_swim_lane(
    mapping = aes(
      x = start_time, xend = end_time, y = pt_id,
      color = disease_assessment
    ),
    linewidth = 5
  ) +
  geom_swim_arrow(
    data = arrow_data,
    mapping = aes(
      xend = end_time,
      y = pt_id,
      arrow = "Continuation"
    ),
    arrow_neck_length = 5,
    arrow_head_length = grid::unit(0.15, "inches")
  ) +
  scale_arrow_discrete(
    limits = "Continuation",
    colours = "firebrick",
    fills = "gold",
    types = "closed",
    name = NULL
  )
```
  
Here, the "Continuation" value mapped to the arrow aesthetic is used by `scale_arrow_discrete()` to determine the arrow’s visual properties. Because this mapping is treated like any other discrete scale in ggplot2, it can also be displayed as a separate legend entry if desired.

Using `scale_arrow_discrete()` is particularly helpful when:

- arrows should appear as a dedicated legend component
- multiple arrow styles are required
- arrow styling should be managed consistently across plots

For simple cases, specifying arrow parameters directly in `geom_swim_arrow()` may still be sufficient. However, using `scale_arrow_discrete()` provides a more flexible and scalable approach when working with more complex swimmer plots.
