---
title: "Gallery"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dev = "ragg_png",
  dpi = 400,
  fig.width = 12, fig.height = 6
)
```

```{r setup, message=FALSE, warning=FALSE}
library(ggswim)
library(ggplot2)
library(dplyr)
```

## Random Data Sets

In this example, we'll set up a random dataset for reproducibility by separately defining a dataframe for our lanes and our markers.

```{r}
set.seed(123)  
lane_data <- tibble(
  x = 0,
  xend = sample(5:20, 30, replace = TRUE),
  y = factor(rep(1:15, each = 2)),
  colour = sample(c('red', 'blue', 'green', 'yellow', 'purple'), 30, replace = TRUE)
)

set.seed(123)  
marker_data <- tibble(
  x = sample(5:20, 30, replace = TRUE),
  y = factor(rep(1:15, each = 2)),
  label = sample(c("A", "B", "C", "D", "E"), 30, replace = TRUE),
  glyph = sample(c("😊", "🎉", "✅", "💥", "✨"), 30, replace = TRUE)
) |> 
  mutate(
    glyph = dplyr::case_when(
      label == "A" ~ "😊",
      label == "B" ~ "🎉",
      label == "C" ~ "✅",
      label == "D" ~ "💥",
      label == "E" ~ "✨",
      .default = NA
    )
  )

lane_data |> 
  rmarkdown::paged_table()
marker_data |> 
  rmarkdown::paged_table()
```

And then we'll call those datasets into their appropriate swim and marker geom functions:

```{r}
ggplot() +
  geom_swim_lane(
    data = lane_data,
    aes(x = x, xend = xend, y = y, colour = colour),
    linewidth = 3
  ) +
  geom_swim_marker(
    data = marker_data,
    aes(x = x, y = y, marker = label),
    size = 8
  ) +
  scale_colour_brewer(name = "Lanes", palette = "Set1") +
  with(marker_data,
       scale_marker_discrete(glyphs = glyph, limits = label, name = "Markers")) +
  labs(title = "Sample Swimmer Plot",
       x = "Time", y = "Record ID") +
  theme_ggswim()
```

## Themeing with ggswim

Here are some example of the themeing functions available in ggswim using the dataset examples from the README:

```{r}
arrow_data <- patient_data |>
  dplyr::left_join(
    end_study_events |>
      dplyr::select(pt_id, label),
    by = "pt_id"
  ) |>
  dplyr::select(pt_id, end_time, label) |>
  dplyr::filter(.by = pt_id, end_time == max(end_time)) |>
  dplyr::filter(is.na(label)) |>
  unique()

all_events <- dplyr::bind_rows(
  infusion_events,
  end_study_events
)

p <- patient_data |>
  ggplot() +
  geom_swim_lane(
    mapping = aes(
      x = start_time, y = pt_id, xend = end_time,
      colour = disease_assessment
    )
  ) +
  geom_swim_arrow(
    data = arrow_data,
    mapping = aes(xend = end_time, y = pt_id),
    arrow_colour = "red",
    arrow_fill = "white"
  ) +
  geom_swim_marker(
    data = all_events,
    aes(x = time_from_initial_infusion, y = pt_id, marker = label),
    size = 5
  ) +
  scale_colour_brewer(
    name = "Disease Assessments",
    palette = "Set1"
  ) +
  scale_marker_discrete(
    glyphs = all_events$glyph,
    colours = all_events$colour,
    limits = all_events$label,
    name = "Study Events"
  ) +
  labs(title = "Clinical Trial Swimmer Plot", subtitle = "Example Swimmer Plot") +
  xlab("Time Since Infusion (Months)") +
  ylab("Patient ID")

p
```

### `theme_ggswim()`

```{r}
p + theme_ggswim()
```

### `theme_ggswim_dark()`

```{r}
p + theme_ggswim_dark()
```
