---
title: "Gallery"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dev = "ragg_png",
  dpi = 400,
  fig.width = 12, fig.height = 6
)
```

```{r setup, message=FALSE, warning=FALSE}
library(ggswim)
library(ggplot2)
library(dplyr)
```

This gallery will showcase different datasets and ways to interact with ggswim.

## Single-Row Dataset

Some common swimmer plot data structures may involve single-row observations where each record or lane corresponds to a single status classification and contains all associated points of interest. Let's look at an example of what we mean by this:

```{r single row dataset}
single_row_data <- tibble::tribble(
  ~"record", ~"status", ~"start_time", ~"end_time", ~"m1", ~"m1_time", ~"m2", ~"m2_time", ~"m2_name",
  1, "status1", 0, 5, "m1", 3, "❌", 5, "Negative",
  2, "status1", -2, 7, "m1", 4, "✅", 6, "Positive",
  3, "status2", 2, 15, "m2", 10, "❌", 15, "Negative"
)

# Convert to factor
single_row_data$record <- factor(single_row_data$record)

single_row_data |>
  rmarkdown::paged_table()
```

In this dataset, we have a combination of statuses and markers where each `record` is a single row observation. Notice that we have markers defined as both points and label value/name pairs. 

For this example, we'll add two additional new scales to separate out the marker points and the marker labels. Notice how we must still refer to `data` to `single_row_data` due to the scale splits:

```{r}
ggplot2::ggplot() +
  geom_swim_lane(
    data = single_row_data,
    mapping = aes(x = start_time, xend = end_time, y = record, colour = status), linewidth = 2
  ) +
  scale_color_manual(name = "Lanes", values = c("firebrick", "steelblue")) +
  new_scale_color() +
  geom_swim_point(
    data = single_row_data,
    mapping = aes(x = m1_time, y = record, colour = m1),
    size = 5
  ) +
  scale_color_manual(name = "Marker Points", values = c("green", "orange")) +
  new_scale_color() +
  geom_swim_label(
    data = single_row_data,
    mapping = aes(x = m2_time, y = record, label_vals = m2, label_names = m2_name),
    label.size = NA, size = 5, fill = NA
  ) +
  scale_color_discrete(name = "Marker Labels") +
  theme_minimal()
```
