---
title: "Gallery"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(ggswim)
library(dplyr)
```

This gallery will explore and showcase different ways to use ggswim.

## Data Spotlight: `pbcseq`

For the next few examples, we will work with the `survival::pbcseq` dataset from the survival package. This dataset was chosen as it contains a number of the elements we might expect to see in survival data including follow-up times, event markers, and subject statuses.

> Consult `?survival::pbcseq` for more information about the dataset.

Let's first take a look at this dataset:

```{r pbcseq dataset}
library(survival)

pbcseq |>
  rmarkdown::paged_table()
```

For this demo, let's pay particular attention to the following variables:

- `id`: This will be our record identifier and make up our y-axis
- `day`: This will be our time series data and make up our x-axis
- `trt`: We'll use this as our "treatment" group to define our fill layer for the lanes
- `status`: A record endpoint indicator, we'll use this to delineate if records continued on or deceased
- `ascites`, `hepato`, and `spiders`: Various binary events that we'll designate as our swimmer plot markers

To make the best use of the space. this swimmer plot will only use the first 20 records, where the lanes stretch the length of the record's enrollment period along the x-axis, are identified by their ID on the y-axis, and filled and grouped by their treatment category (`trt`).

Some pre-work is necessary to get our data in the right format and shape for a swimmer plot. Most notably, `day` is not quite ready for the bar plot that makes up the lanes of ggswim. For this dataset, the start date is always 0, but had it been variable then we might have had to adjust our code to encompass a start and end date (see [Plotting with Negative Swimmer Lanes](#plotting-with-negative-swimmer-lanes) for an example). Fortunately, we only really need the max date of each record's enrollment. Lastly, converting some of these variables to factors will help us with arranging the swim plot in the most effective way.

```{r pbcseq preparation}
pbcseq_prep <- survival::pbcseq |>
  filter(id <= 20) |>
  mutate(
    .by = id,
    min_day = min(day),
    max_day = max(day)
  ) |>
  relocate(c(min_day, max_day), .after = day) |>
  arrange(trt, max_day) |>
  mutate(id = factor(id, levels = unique(id)),
         trt = factor(trt))

pbcseq_prep |>
  rmarkdown::paged_table()
```

Now all we need is to subset the original data for what will comprise our swimmer lanes. We'll also add a column called `alive`, which will be `TRUE` if the patient was not declared deceased at the end of the enrollment period (`status == 2`). This will allow us to add arrows onto our plot later (see [Adding arrows to ggswim](articles/add_arrows.html) for more information).

```{r pbcseq lanes}
pbcseq_lanes <- pbcseq_prep |>
  select(id, trt, max_day, status) |>
  mutate(alive = ifelse(status == "2", FALSE, TRUE)) |>
  unique()

pbcseq_lanes |>
  rmarkdown::paged_table()
```

To prepare a swimmer markers data subset, we'll focus just on `ascites`, `hepato`, and `spiders` in this example. A `pivot_longer()` from the tidyr package lets us expand this subset into an easily manageable dataframe containing the marker type, the day that marker occurred, and the associated record.

```{r pbcseq markers}
pbcseq_markers <- pbcseq_prep |>
  select(id, day, ascites, hepato, spiders) |>
  tidyr::pivot_longer(c(ascites, hepato, spiders),
                      names_to = "marker") |>
  filter(value == 1)

pbcseq_markers |>
  rmarkdown::paged_table()
```

Now we have everything we need to make a swimmer plot! Our first call to `ggswim()` makes use of `pbseq_lanes` dataframe and our single call to `add_marker()` uses the newly created `pbseq_markers` dataframe. It is important to note that we specify `trt` as the **fill** option for the lanes and `marker` as the **color** for the markers. These aesthetic parameters are exclusive to these functions, and are explained in greater detail in the documentation. 

```{r pbcseq swim plot}
ggswim(
  data = pbcseq_lanes, aes(x = max_day, y = id, fill = trt),
  arrow = "alive"
) +
  add_marker(
    data = pbcseq_markers,
    aes(x = day, y = id, color = marker, shape = marker),
    size = 3, alpha = 0.5
  ) +
  theme_ggswim()
```

## Plotting with Negative Swimmer Lanes

Let's say we have a dataset where our analysis period has varying starting time points, some of which occur before 0. Up until now, we've only ever really worked with classic datasets that look at 0 as the initial time point and move positively on the x-axis. To show this, we'll modify the internal `patient_status` dataframe into a new dataframe called `patient_status_negative`:

```{r patient_status_negative}
set.seed(135)
patient_status_negative <- patient_status |>
  mutate(
    value = case_when(
      value == 0 ~ sample(c(-5:0), size = 30, replace = TRUE),
      TRUE ~ value
    )
  )

patient_status_negative |>
  rmarkdown::paged_table()
```

Notice how we replaced any initial starting point for a given record of 0 with a random value between -5 and 0. Now let's plot with `ggswim()`:

```{r negative ggswim}
ggswim(
  patient_status_negative,
  aes(x = value, y = subject_id, fill = cohort)
)
```

ggswim is able to handle any starting point, even if those starting points are variable. It just requires a little extra work to get the data in the right format, such as pivoting to get the time series data under oen column.

This plot looks good, but let's make it a little more aligned with how we would want to display the data. Let's clean it a bit and give it an identifiable vertical line to separate positive and negative time periods:

```{r negative swim pretty}
ggswim(
  patient_status_negative,
  aes(x = value, y = subject_id, fill = cohort)
) +
  ggplot2::geom_vline(xintercept = 0, linetype = 3) +
  theme_ggswim()
```
