---
title: "Getting Started with ggswim"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 12,
  fig.height = 6
)
```

The ggswim package eases the development of swimmer plots in R through integration with the ggplot2 framework. In this vignette, we'll walk through how users can create visually striking swimmer plots.

At the heart of ggswim are two core functions: `ggswim()` and `add_marker()`. The former, an extension of `geom_col()`, allows users to construct the horizontal bars, what we'll sometimes refer to as "lanes." Meanwhile, `add_marker()` leverages `geom_point()` and `geom_label()` to effortlessly embed key events, or "markers," onto the lanes. These can take the form of unicode shapes, symbols, or even emojis.

Drawing from the well-established principles of ggplot2, ggswim allows users to apply familiar layer-building techniques, including the application of styles and themes.

### Building swimmer plots with aesthetic mapping

To start, let's build a swimmer plot using `aes()` and dynamically mapped data (for static mapping, see [Static Mapping with Markers](https://chop-cgtinformatics.github.io/ggswim/articles/add_arrows.html)). As with the README, we will be using ggswim's internal datasets: `patient_data`, `infusion_events`, and `end_study_events`. 

Let's start with observing `patient_data`'s structure:

```{r patient_data , echo=FALSE, include=TRUE}
library(ggswim)

patient_data |>
  rmarkdown::paged_table()
```

`patient_data` contains a long pivoted dataset where patient ID's (`pt_id`) can be repeated. These rows are differentiated by `event_marker`s combined with corresponding times where `delta_t0` is the change in time (in days) from an infusion. Together, these rows signify a patient timeline with consideration for infusions, disease assessments, and B-cell status changes.

- `disease_assessment_status`: The overall status at a disease assessment coupled with B-cell status changes where:
   - CR = "Complete Response"
   - CRi = "Complete Response with Incomplete Blood Count Recovery"
   - RD = "Relapsed Disease"

Now, let's make the plot using `ggswim()` and our first `add_marker()` call:

```{r ggswim plot}
p <- ggswim(
  patient_data,
  mapping = aes(
    x = delta_t0_months,
    y = pt_id, fill = disease_assessment
  )
)

p
```

Here we have a simple bar chart, showing infusions grouped by patients with a given disease assessment status. `ggswim()` does the work of setting up a `geom_col()` and readying our plot layers for additional ggswim-specific features such as the "markers" mentioned earlier. Let's add a marker layer next from the `infusion_events` dataset:

```{r infusion_events}
infusion_events |>
  rmarkdown::paged_table()
```

This dataset is much simpler, having only 3 columns indicating infusions (considered `t0`) and reinfusions at some point beyond `t0` (if a patient had any).

```{r first marker plot, message=FALSE}
p <- p +
  add_marker(
    data = infusion_events,
    aes(
      x = delta_t0_months,
      y = pt_id,
      name = "Infusion"
    ),
    size = 5, position = "identity", alpha = .75, color = "green"
  )

p
```

Now we can see shapes have been added to establish our markers for infusions.

Our last dataset involves end of study events, i.e. events that indicate a patient has left the study for various reasons.

```{r end_study_events}
end_study_events |>
  rmarkdown::paged_table()
```

You'll notice that this dataset includes use of emojis under `end_study_label`. In addition to shapes and symbols, ggswim supports the use of emojis when rendering swimmer plots. `add_marker()` allows users to specify shapes or emojis by using the appropriate `aes()` argument callouts: `label_vals` and `label_names`. These are unique mapping parameters to ggswim. Let's add a layer using emojis via the `end_study_events` dataset:

```{r second marker plot, message=FALSE, warning=FALSE}
p <- p +
  add_marker(
    data = end_study_events,
    aes(
      x = delta_t0_months,
      y = pt_id,
      label_vals = end_study_label,
      label_names = end_study_name
    ),
    label.size = NA, fill = NA, size = 5
  )

p
```

We've successfully made a swimmer plot with lanes and two different kinds of marker layers. Recall that ggswim works within the ggplot2 framework, therefore customization can be done using the same ggplot2 theme and styling functions users may already be familiar with. Let's fix up this plot so it looks nicer on the eyes:

```{r mapping finalizing, message=FALSE, warning=FALSE}
library(ggplot2)

p <- p +
  theme_minimal() +
  scale_colour_manual(
    values = c("firebrick", "#F5EB0A", "gray50", "#25DA6D", "#25DA6D", NA, NA, NA)
  ) +
  scale_shape_manual(
    values = c(19, 19, 15, 17, 18)
  ) +
  scale_fill_manual(
    name = "Overall Disease Assessment",
    values = c("#6394F3", "#F3C363", "#EB792F")
  ) +
  labs(title = "My Swimmer Plot") +
  xlab("Time (Months)") + ylab("Patient ID")

p
```

We can also supply the `theme_ggswim()` function:

```{r mapping with theme_ggswim, message=FALSE, warning=FALSE}
p +
  theme_ggswim()
```

### Additional notes

Some additional considerations to keep in mind when working with ggswim:

- **Handling Missing Data:** ggswim does not support missing data for mapping aesthetics. If any are detected, developers will receive a warning, and the missing data may appear as `NA` values in the display, but will be excluded from the legend.

- **Color Mapping with `ggswim()`:** Please note that `ggswim()` does not support mapping using the `colour` argument.

- **Filling Markers with `add_marker()`:** Similarly, `add_marker()` does not support mapping using the `fill` argument.

- **Rendering Emojis and Custom Shapes:** To ensure emojis and other custom shapes display correctly, users may need to [switch their graphics rendering device](https://support.posit.co/hc/en-us/articles/360017886674-Troubleshooting-RStudio-Rendering-Errors) to AGG.
