---
title: "Getting Started with ggswim"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The ggswim package eases the development of swimmer plots in R by seamlessly integrating with the ggplot2 framework. In this vignette, we'll walk through how users can create visually striking swimmer plots.

At the heart of ggswim are two pivotal functions: `ggswim()` and `add_marker()`. The former, an extension of `geom_col()`, allows users to construct the fundamental horizontal bars, often referred to as "lanes." Meanwhile, `add_marker()` leverages `geom_point()` and `geom_label()` to effortlessly embed key events or "markers" onto the plot. These can take the form of unicode shapes, symbols, or even emojis.

Drawing from the well-established principles of ggplot2, ggswim allows users to apply familiar layer-building techniques, including the application of styles and themes.

Let's dive in and explore the boundless possibilities that `ggswim` offers for expressive swimmer plots!

### Building swimmer plots with aesthetic mapping

To start, let's build a swimmer plot using `aes()` and dynamic mapping of data. As with the README, we will be using ggswim's internal `patient_data` list dataset. 

As mentioned in the README, `patient_status` is built to reflect a typical dataset in the wild. To make this work well and appear nicely in our swimmer plot output, a little bit of preparation was required. Unlike what we provide for you out of the box, a more typical `patient_status` might look something like this:

```{r load dplyr, include=FALSE}
library(dplyr)
```

```{r prep patient_status, echo=FALSE, include=TRUE}
set.seed(123)

tibble::tibble(subject_id = rep(1:10, length.out = 15)) |>
  sample_n(15) |>
  arrange(subject_id) |>
  mutate(
    .by = subject_id,
    cohort = rep(sample(c("Cohort A", "Cohort B"), 1), length.out = n(), each = length(unique(subject_id))),
    cohort = factor(cohort, levels = c("Cohort B", "Cohort A")),
    status = ifelse(duplicated(subject_id) | !duplicated(subject_id, fromLast = TRUE), "On Study", "Off Study"),
    status = factor(status, levels = c("On Study", "Off Study"), ordered = TRUE),
    time_end = ifelse(status == "On Study", runif(n(), 1, 10), runif(n(), 11, 20)),
    time_start = ifelse(status == "On Study", 0, NA),
    time_start = ifelse(is.na(time_start), lead(time_end), time_start)
  ) |>
  arrange(subject_id, status) |>
  relocate(time_start, .before = time_end) |>
  mutate(subject_id = factor(subject_id))
```

Therefore, when viewing the available `patient_status`, keep in mind it has undergone the following data cleaning steps:

- A pivot that keeps `time_start` and `time_end` under one column
- The addition of a sorting variable, `time_sorting`, to support arrangement using summed times
- Arrangement of the data by cohort groups, time sorted values, and subject ID factor levels

```{r}
library(ggswim)

patient_status
```

Now, let's make the plot using `ggswim()` and our first `add_marker()` call:

```{r aes mapping plot}
p <- ggswim(
  data = patient_status,
  mapping = aes(
    x = value,
    y = subject_id,
    fill = cohort
  )
) +
  add_marker(
    data = adverse_events,
    mapping = aes(
      x = time_of_event,
      y = subject_id,
      color = adverse_event_name,
      shape = adverse_event_name
    ),
    size = 5
  )

p
```

In addition to shapes and symbols, ggswim supports the use of emojis when rendering swimmer plots. `add_marker()` allows users to specify shapes or emojis by using the appropriate argument callouts (`label` and `shape`). Let's add a layer using emojis via the `medication_administration` dataset:

```{r label mapping}
p <- p +
  add_marker(
    data = medication_administration,
    mapping = aes(
      x = time_of_event,
      y = subject_id,
      label = medication,
      color = name
    ),
    label.size = NA, fill = NA, size = 5
  )

p
```

At this point we have a working swimmer plot with lanes and two different kinds of marker layers. Recall that ggswim works within the ggplot2 framework, therefore customization can be done using the same ggplot2 theme and styling functions users may already be familiar with:

```{r mapping finalizing, message=FALSE, warning=FALSE}
library(ggplot2)

p +
  labs(x = "Time", y = "Subject ID", color = "Markers") +
  ggtitle("My Swim Plot") +
  theme_minimal() +
  scale_color_manual(
    name = "Markers",
    values = c("firebrick", "tomato", NA, NA, "chartreuse2")
  ) +
  scale_shape_manual(
    name = "Markers",
    values = c(19, 15, 8, 18)
  ) +
  scale_fill_manual(
    name = "Lanes",
    values = c("steelblue", "cyan")
  )
```

However, ggswim also provides a convenient theme function all of its own if desired:

```{r mapping with theme_ggswim}

p +
  labs(x = "Time", y = "Subject ID", color = "Markers") +
  ggtitle("My Swim Plot") +
  theme_ggswim()
```

### Customizing Markers: A Deeper Dive

In the world of data, one size rarely fits all. That's why ggswim caters to both dynamic and static marker assignments. But what does that mean exactly?

Imagine you have a dataset where you want to manually define the color of a single, significant marker rather than relying on a mapped variable. By splitting up `adverse_events` into distinct data frames, you can achieve this customized effect.

To make it all come together, make sure to include an `aes()` `name` argument within `add_marker()`. This serves as a crucial anchor point for the legend, ensuring clarity in your visualization.


```{r static mapping}
adverse_event_cardio <- adverse_events |>
  dplyr::filter(adverse_event_name == "Cardiac Disorder")

adverse_event_inf <- adverse_events |>
  dplyr::filter(adverse_event_name == "Infection")

adverse_event_psych <- adverse_events |>
  dplyr::filter(adverse_event_name == "Psychiatric Disorder")
```

Now, let's plot it similarly to how we have previously:

```{r static plot, message=FALSE, warning=FALSE}
ggswim(
  data = patient_status,
  mapping = aes(
    x = value,
    y = subject_id,
    fill = cohort
  )
) +
  add_marker(
    data = adverse_event_cardio,
    mapping = aes(
      x = time_of_event,
      y = subject_id,
      name = "AE: Cardiac Disorder"
    ),
    color = "red", size = 5
  ) +
  add_marker(
    data = adverse_event_inf,
    mapping = aes(
      x = time_of_event,
      y = subject_id,
      name = "AE: Infection"
    ),
    color = "blue", size = 5, shape = 12
  ) +
  add_marker(
    data = adverse_event_psych,
    mapping = aes(
      x = time_of_event,
      y = subject_id,
      name = "AE: Psychiatric Disorder"
    ),
    color = "green", size = 5, shape = 2
  ) +
  theme_ggswim()
```

This example showcases ggswim's ability to handle different data structures from varying sources. Its adaptable framework lets users customize the output precisely to their unique requirements.

### Additional notes

Here are some essential points to keep in mind when working with ggswim:

- **Handling Missing Data:** ggswim does not support missing data for mapping aesthetics. If any are detected, developers will receive a warning, and the missing data may appear as `NA` values in the display, but will be excluded from the legend.

- **Color Mapping with `ggswim()`:** Please note that `ggswim()` does not support mapping using the `colour` argument.

- **Filling Markers with `add_marker()`:** Similarly, `add_marker()` does not support mapping using the `fill` argument.

- **Rendering Emojis and Custom Shapes:** To ensure emojis and other custom shapes display correctly, users may need to switch the graphics rendering device to AGG.

These considerations will help you make the most of ggswim while building your visualizations.
